An Interpretable Machine Learning Framework for Modeling High-Resolution Spectroscopic Data\footnote{Open source code at \url{https://github.com/gully/blase}}

### 2022-10-03 paper/paper1/ms.tex
Grammarly edits and some extra references, etc.

In this paper, we show that autodiff-aware semi-empirical models offer an appealing middle ground: informed by self-consistent models but refined with data.  This sweet spot in the bias-variance tradoff can be thought of as a hybrid data-and-model driven approach.  The algorithm presented here focuses on modeling the spectra of stars and brown dwarfs.  Existing stellar models \citep[\emph{e.g.}][]{husser13} and substellar models \citep[\emph{e.g.}][]{2021ApJ...920...85M} laboriously solve for a self consistent thermal structure in the atmosphere given the copious opacity sources that themselves depend on temperture and pressure.  Here we build upon that hard work by cloning pre-existing synthetic stellar or substellar models (Section \ref{methodology}), and optionally by cloning models of Earth's atmospheric ``telluric'' absorption (Section \ref{sectionTelluric}). We introduce the interpretable forward-model design and its \texttt{PyTorch}-based \citep{2019arXiv191201703P} implementation, \texttt{blase}.  In Section \ref{transferLearn}  we describe how to adapt both stellar and telluric cloned models simultaneously, using a transfer-learning step.  We obtain semi-empirical models by comparing to real-world \'echelle data in Section \ref{secResults}. Finally, we discuss perspectives on how to think of \emph{blas\'e} (Section \ref{secDiscuss}) and chronicle many conceivable extensions for unlocking new science (Section \ref{secFutureWork}).

@@

We truncate the red and blue limits of the precomputed synthetic spectrum to match a high-bandwidth echelle spectrograph, extended with a buffer at the edges of size $\pm \Delta \lambda_{\mathrm{buffer}}$.  \edit1{The buffers are needed for two reasons: 1) to include the influence of spectral lines whose centers lie just outside the region-of-interest, but whose wings are broad enough to affect the neighboring pixels; and 2) to account for astrophysical radial velocity shifts that may send line centers into- and out-of- the region of interest.}  We set the value of the buffers to account for plausible radial velocity and rotational broadening of real stars. A generous buffer of $v \sin{i} < 500 \;\kmps$ and $|RV|<500 \;\kmps$ yields a typical buffer of about 30~\AA.

@@

We initialize the cloned model with a series of preprocessing steps. We divide the entire spectrum by a black body $\mathsf{B}(\bm{\lambda}_S)$ of the same effective temperature $T_{\mathrm{eff}}$ as the model template. The resulting signal usually still has smooth wiggles around the continuum. An optional continuum flattening step ensures that subsequent spectral line finding steps get applied uniformly. This high-pass filtering step should be set to capture the genuine spectral shape, without over-fitting broad line wings such as those in deep hydrogen and sodium lines. Any high-pass filter will work, a Gaussian Process approach would be ideal \citep{czekala15}.  Instead we apply a simple and familiar heuristic: fit a polynomial $\mathsf{P}(\bm{\lambda}_S)$ to a few continuum peaks and divide out the trend.

@@

\noindent where the division indicates element-wise division of these arrays or ``vectors'' of flux values.  \edit1{For elementwise-multiplication, we will adopt the $\odot$ operator symbol to distinguish that the multiplication occurs elementwise between two ``vectors'' of spectral fluxes, as opposed to a dot product (scalar ouput) or matrix multiplication (matrix output).  This style of vector multiplication and division is standard in numerical array manipulations such as those in \texttt{numpy}, \texttt{IDL}, \texttt{matlab}, \emph{etc.}}

@@

A recreation of the unvarnished input spectrum---if desired---can be obtained by multiplying the continuum-flattened signal by the ``perturbed black body'', $\mathsf{B}(\bm{\lambda}_S)\odot \mathsf{P}(\bm{\lambda}_S)$, that symbolizes the black body modulated by continuum opacity or broad-band radiative transfer effects.  This smooth spectrum may be useful for applications that need to keep track of broad-band flux, such as low-resolution spectra, or regions with molecular band heads. The ``perturbed black body'' continuum model contains $n_{\mathrm{poly}}+1$ fixed-but-possibly tunable lookup parameters, plus the fixed input $T_{\mathrm{eff}}$.  For most practitioners these terms serve as nuisance parameters and are perfunctorily discarded.

@@

The Voigt profile $\mathsf{V}(\lambda, \sigma_j, \gamma_j)$ can be computed in exact closed-form using the Voigt-Hjerting function \citep{1938ApJ....88..508H} as the real part of the complex Fadeeva function \citep[\emph{e.g.}][]{2011arXiv1106.0151Z}. Evaluation of the Fadeeva function can be computationally costly, and so approximate forms may be desirable. Here we adopt the pseudo-Voigt approximation \citep{Ida:nt0146}.

@@

\subsection{Goodness of fit metric and gradients}
The model evaluated with its coarse initial values would have terrible performance: it would only vaguely resemble the synthetic spectral model, with up to $\pm 50\%$ undulations from the inexact assignment of widths, lineshapes, and amplitudes. Instead, we tune the parameters of the model, starting from these coarse initial values. This model has between $N_{\mathrm{lines}}\times 1$ and $N_{\mathrm{lines}}\times 4$ free parameters, depending on how many of the 4 physical line properties you wish to tune.  \edit2{The treshold-based line-finding algorithm performs very well at finding the line center positions, so in practice we can fix the line center wavelengths.}  We therefore default to fitting three parameters per line, where the center wavelength is held fixed and the amplitude, width, and lineshape are allowed to vary. We minimize a scalar ``goodness-of-fit'' metric, \emph{aka} loss scalar $\mathcal{L}$, chosen as the mean squared error (MSE), which is proportional to $\chi^2$, the sum of the squares of the residual vector $\mathsf{R} \equiv \mathsf{S}-\mathsf{S}_{\rm clone}$ but has no notion of per-pixel noise since the precomputed synthetic spectrum has no uncertainty:

@@

\edit1{Geometrically, the $\grad \mathcal{L}$ Jacobian resembles a compass needle guiding the direction to the best fit. This compass needle resides in, say, 21,000 dimensional space, instead of our mere 3 dimensions of physical space.  Conventional gradient-free optimizers evaluate the Jacobian through approximate finite differences, which result in numerical instabilities.  Those instabilities accrue and break down at around the 300 dimension mark.  So non-gradient-based optimization can be pictured as having only a coarse needle---assembled from approximate finite differences---that begins to spin aimlessly after too many dimensions.}

@@

It is this ability to automatically compute Jacobians that sets \texttt{PyTorch} (and \texttt{JAX} and \texttt{TensorFlow}) apart.  These frameworks give exact Jacobians, instantaneously, for free (or cheap).  Without exact Jacobians a conventional optimization step only obtains \emph{one} piece of information: how much the overall loss changed.  With exact Jacobians, we obtain $N_\mathrm{params}+1$ pieces of information for each evaluation of the forward model.  The power of gradient descent becomes transformative as the number of parameters grows into the tens of thousands, as is the case for \emph{blas\'e}.

@@

Forward modeling with tens of thousands of physics-informed parameters may seem like such a significant paradigm shift that it can feel too good to be true.  In this section we introduce the non-negligible architectural design tradeoffs that arise when adopting an autodiff framework---such as \texttt{PyTorch}---for physics based forward modeling.

First, we make a few tweaks to the implementation for numerical purposes. We enforce that all Gaussian and Lorenzian widths are positive by tuning the natural log of the widths, then exponentiating them before inclusion in Equation \ref{equation1}. Based on the minus sign in Equation \ref{equation1}, an amplitude $a_j$ could hypothetically take on either positive values (flux loss, absorption) or negative values (flux gain, emission).  For now we focus on photospheric absorption lines---as opposed to, say, chromospheric emission lines---and therefore enforce all the amplitudes to be positive by tuning $\ln{a_j}$ and then exponentiating in the same strategy as above.  Emission lines---if desired---could be included with a mere sign flip to Equation \ref{equation1}.  Lines that can manifest either in absorption or emission could hypothetically relax the natural log pre-processing step for isolated lines.  We narate only the absorption scenario moving forward.

@@

Sparse matrix methods generally support an operation known as \emph{coalescing}, which sums values with repeated indices. Each pixel may get computed about $\sim100$ times in this sparse implementation, which is about $70\times$ faster than each pixel getting computed $N_\mathrm{lines}\sim7000$ times in the dense approach.  Efficient algorithms for assembling and coalescing sparse matrices exist in \texttt{PyTorch}.  Some GPUs now support additional hardware-acceleration of sparse matrices, providing even greater speedups.

@@

\begin{figure}[hbt!]
    \centering
    \includegraphics[width=1.0\columnwidth]{blase_clone_demo.png}
    \caption{PHOENIX spectrum cloned with \emph{blas\'e}.  This $T_{\mathrm{eff}}=4700\;$K,  $\log{g}=4.5$ solar metallicity model has 9,028 individual cloned spectral lines, each with 3 tuned parameters.  The pictured 50 \AA~ chunk contains 121 spectral lines and represents about 1$\%$ of the entire spectral bandwidth that was cloned.  Some flaws can be seen near the cores of deep lines, or wings of broad lines.}
    \label{fig_cloned_spectrum_demo}
\end{figure}

@@

\begin{figure*}[hbt!]
    \centering
    \includegraphics[width=0.98\textwidth]{TelFit_clone_2panel.pdf}
    \caption{Pre-computed TelFit model cloned with \texttt{blase}.  \textbf{Before:} The top panel shows a forward model evaluated with coarsely initialized line-by-line properties, based on simple threshold-based peak-finding and crude estimation of the shape of the line.  \textbf{After:} The bottom panel shows the \texttt{blase} fine-tuned clone, with nearly indentical spectral structure.}
    \label{fig_telluric_clone}
\end{figure*}

@@

We compute the residual $\mathsf{R}(\bm{\lambda}_S)$ of native-minus-cloned PHOENIX model, illustrated in the bottom panel of Figure \ref{fig_cloned_spectrum_demo}. We see an RMS residual of 1.2\%/pixel at native resolution.  The telluric clone shows a comparable level of performance.  These residuals tend to pile up in local symmetrically balanced clusters that get cancelled out once convolved with coarser instrumental line profiles, so their overall effect at instrumental resolution is typically negligible: the clones are almost perfect.  We identify three main categories of cloning flaws that may not be negligible depending on the science application.

@@

Second, a conspicuous flaw occurs in the line cores of relatively narrow lines.  The cloned model tends to overestimate the flux at the core and underestimate the flux along the slopes of the lines.  \edit1{The residuals therefore exhibit a \texttt{W}-shape ringing artifact.  The scientific impact of this flaw will depend on the precision demands of the application, but it may not be as bad as it seems for many applications.  Recall that the cloning occurs at native spectral resolution, much greater than typical instrumental resolution.  The \texttt{W}-shape ringing residuals will tend to cancel each other out once the clone model undergoes convolution with an instrumental profile (Section \ref{jointModel}).  We ascribe this flaw to the adoption of approximate pseudo-Voigt profiles rather than exact Voigt profiles.  High-precision applications may wish to employ the exact Voigt profile, which we estimate to cost about 27$\times$ more than this approximate form.}

\edit1{The overprediction of line cores in narrow lines is only partly attributable to the pseudo-Voigt approximation.  Even exact Voigt profiles inadequately represent true atmospheric opacities. Typical stellar and substellar atmosphere models compute opacities for a stratified atmosphere consisting of tens or hundreds of locally isothermal, isobaric atmospheric layers.  Each of those layers may exhibit a local per-line opacity that matches a Voigt profile.  But the cumulative product of many slightly-different Voigt profiles is not exactly equal to any single Voigt profile.  So the \emph{blas\'e} approach of asserting lines as Voigt profiles can be best understood as a phenomenological model-- that lines tend to look Voigt-like, even if they are non-Voigt in detail.  In principle, \texttt{blas\'e} could be configured in a way to learn the requisite perturbations to the Voigt profile in a systematic manner, pooling information among lines.  Such an approach is not yet implemented, and may be an enticing avenue of future research.}

@@

\begin{figure}[hbt!]
    \centering
    \includegraphics[width=0.48\textwidth]{blase_clone_demo_zoom.png}
    \caption{Zoom in of the region between the vertical gray bars in Figure \ref{fig_cloned_spectrum_demo}. The cloned model has 7 spectral lines describing the 400 pixels in this 4 \AA~ chunk.  The native PHOENIX pixel sampling can be see as the boxy steps in both the native and cloned model.  }
    \label{fig_zoom_cloning_performance}
\end{figure}

@@

The cloned models already represents a useful intermediate product: the distillation of $N_s=335,849\times2$ pixel flux values and their wavelength coordinates into a more compact quartet of properties for a list of $<10,000$ spectral lines, a dimensionality reduction of $18\times$ for the cost of 1.2\%/pixel in accuracy.  So the process so far can be myopically viewed as a physics-informed compression algorithm.  But the cloned model serves as a mere stepping stone in our principal quest: the comparison of models to real data.

\subsection{Augmenting the stellar clone with radial velocity and rotational broadening }
Real stars possess two key extrinsic properties.  Rotational broadening $v\sin{i}$ and radial velocity $RV$ depend on the observer's viewing location. We follow \citet{czekala15} by emphasizing the qualifier \emph{extrinsic}, to distiguish between stellar \emph{intrinsic} properties, such as $T_{\mathrm{eff}}, \log{g},\mathrm{and\,} [\mathrm{Fe}/\mathrm{H}]$.  Intrinsic properties appear the same from any viewing location---at least for stars with isotropic surfaces---while extrinsic properties do not.  The distinction is important because the extrinsic terms act as simple convolutions and translations to the cloned spectrum, and can be treated after the cloning procedure.  We therefore define an augmented model, which we designate the ``extrinsic'' model, $\mathsf{S}_{\rm ext}$:

@@

The $\mathsf{S}_{\rm ext}$ spectrum is shown in Step 2 of Figure \ref{blase_flowchart}.  There is currently no equivalent post-processing of the telluric spectrum $\mathsf{T}$.  As mentioned previously, we assume motions of the Earth's atmosphere are much less than the desired stellar radial velocity precisions.  However, demanding Extreme Precision Radial Velocity (EPRV) exoplanet searches may need to consider minuscule systematic $RV$ shifts and broadening of the telluric templates, arising from the turbulent and bulk motions of the Earth's atmosphere.  A $\mathsf{T}_{\rm ext}$ could hypothetically be included in  \emph{blas\'e} to achieve these strenuous precision demands.

@@

Notice that the order of Steps 3 and 4 in Figure \ref{blase_flowchart} cannot be swapped.  Mathematically speaking, elementwise-multiplication and convolution do not commute.  While the distinction may seem negligible, it matters at the level of precisions sought in EPRV applications (Suvrath Mahadevan, Arpita Roy, Sharon Xuesong Wang \emph{priv. comm.}).  Water vapor lines in our own atmosphere can ``beat'' with water vapor in the spectrum of, say, an M-dwarf atmosphere.  The systematic telluric mis-cancelation would imbue a Moir\'e pattern of residuals that is most acute for sources with sharp lines, namely low projected rotational broadening ($v\sin{i}\sim\frac{c}{2R}$).  The approach in \emph{blas\'e} may therefore unlock a level of telluric calibration that has evaded previous efforts.

As noted, typical data-pixel sampling $\bm{\lambda}_D$ is much coarser than the model pixel sampling $\bm{\lambda}_Z$.  We therefore resample the model to the data spectrum in the following way.  We evaluate the joint model $\mathsf{M}_{\rm joint}$ at all of the super-resolution wavelength coordinates, and then compute the mean value of those pixels within the bounds of each coarse data pixel.  The resampling procedure is autodiff-aware: the same clusters of high resolution coordinates map to the same data pixel coordinates, no matter what the $RV$ is.  The $RV$ only dictates what flux values are realized within those pixel bounds.

@@

It may appear desirable to simply optimize all of these parameters in a \emph{laissez-faire} manner, allowing them all to take on whatever value the data dictates.  Such a stratagem would overfit the data, resulting in unphysically perverse lineshapes that do not reflect the air of reality we aspire to impose on our synthetic spectral models.  Lines would haphazardly fit noise-spikes, and conspire together to warp spectral shapes in unexpected ways.  This overfit model may suit some rare purposes.  But most of the time, we prefer to strike a better balance in the bias-variance tradeoff.

@@

The degree of regularization will control the extent of overfitting or underfitting. The most extreme regularization---the antithesis of the \emph{laissez-faire} scenario---would yield a model too rigid to respond to the data at all, yielding a model entirely unchanged from the cloned PHOENIX and \texttt{TelFit} models, the extreme end of underfitting.  So regularization constitutes the only hyperparameters worthy of tuning in \emph{blas\'e}.  The choice of how to set the regularization is problem-specific.  We default to the following choice. We fix all line parameters except for amplitude, which receives an L1 loss-- namely we penalize the absolute value of departures from an amplitude's starting place:

@@

Figure \ref{fig_WASP69_demo} shows a portion of an HPF spectrum of WASP 69 centered on a region devoid of telluric lines.  This figure highlights a baseline case with no line-by-line fitting, simply conventional template matching: comparing HPF data of WASP 69 to the closest PHOENIX template, convolved and resampled to the HPF resolution and sampling.  Conspicuous residuals of $\pm10\%$ appear throughout the spectrum, with lines in the correct place, but with the amplitudes systematically biased.  These lines residuals arise from bona-fide imperfections in the PHOENIX spectrum, with a minor contribution from the coarse sampling of the PHOENIX grid.

Figure \ref{fig_WASP69_transferred} shows the same data spectrum compared to a pixel-level model trained using the \emph{blas\'e} technique described in Section \ref{transferLearn}.  The model fit appears much better, with typical residuals approaching the photon noise of the data themselves.  The model is not perfect however, especially around line cores.  These residuals stem from a combination of causes.  First, our finite regularization restricts line amplitudes and widths from straying too far from their values.  This computational tug-of-war makes line cores land just-short of the values the otherwise would have obtained in the absence of regularization.  Second, the spectral resolution kernel may get biased from the domineering model-misspecification of broad lines, setting up a slightly sub-par performance for all the other lines.

@@

Up to this point we have presented the mechanics of  \emph{blas\'e}, and have shown that it applies to a wide range of applications including stars, substars, and Earth's atmosphere. The interpretation of the approach and its line-by-line outputs may carry different meaning to different practitioners.  Here we caricature some of these perspectives:

>>>>

In this paper, we show that autodiff-aware semi-empirical models offer an appealing middle ground: informed by self-consistent models but refined with data.  This sweet spot in the bias-variance tradeoff can be thought of as a hybrid data-and-model driven approach.  The algorithm presented here focuses on modeling the spectra of stars and brown dwarfs.  Existing stellar models \citep[\emph{e.g.}][]{husser13} and substellar models \citep[\emph{e.g.}][]{2021ApJ...920...85M} laboriously solve for a self-consistent thermal structure in the atmosphere given the copious opacity sources that themselves depend on temperature and pressure.  Here we build upon that hard work by cloning pre-existing synthetic stellar or substellar models (Section \ref{methodology}), and optionally by cloning models of Earth's atmospheric ``telluric'' absorption (Section \ref{sectionTelluric}). We introduce the interpretable forward-model design and its \texttt{PyTorch}-based \citep{2019arXiv191201703P} implementation, \texttt{blase}.  In Section \ref{transferLearn}  we describe how to adapt both stellar and telluric cloned models simultaneously, using a transfer-learning step.  We obtain semi-empirical models by comparing to real-world \'echelle data in Section \ref{secResults}. Finally, we discuss perspectives on how to think of \emph{blas\'e} (Section \ref{secDiscuss}) and chronicle many conceivable extensions for unlocking new science (Section \ref{secFutureWork}).

@@

We truncate the red and blue limits of the precomputed synthetic spectrum to match a high-bandwidth echelle spectrograph, extended with a buffer at the edges of size $\pm \Delta \lambda_{\mathrm{buffer}}$.  \edit1{The buffers are needed for two reasons: 1) to include the influence of spectral lines whose centers lie just outside the region-of-interest, but whose wings are broad enough to affect the neighboring pixels; and 2) to account for astrophysical radial velocity shifts that may send line centers into- and out of- the region of interest.}  We set the value of the buffers to account for plausible radial velocity and rotational broadening of real stars. A generous buffer of $v \sin{i} < 500 \;\kmps$ and $|RV|<500 \;\kmps$ yields a typical buffer of about 30~\AA.

@@

We initialize the cloned model with a series of preprocessing steps. We divide the entire spectrum by a black body $\mathsf{B}(\bm{\lambda}_S)$ of the same effective temperature $T_{\mathrm{eff}}$ as the model template. The resulting signal usually still has smooth wiggles around the continuum. An optional continuum flattening step ensures that subsequent spectral line finding steps get applied uniformly. This high-pass filtering step should be set to capture the genuine spectral shape, without over-fitting broad line wings such as those in deep hydrogen and sodium lines. Any high-pass filter will work, a Gaussian Process approach would be ideal \citep{czekala15}.  Instead, we apply a simple and familiar heuristic: fit a polynomial $\mathsf{P}(\bm{\lambda}_S)$ to a few continuum peaks and divide out the trend.

@@

\noindent where the division indicates element-wise division of these arrays or ``vectors'' of flux values.  \edit1{For elementwise-multiplication, we will adopt the $\odot$ operator symbol to distinguish that the multiplication occurs elementwise between two ``vectors'' of spectral fluxes, as opposed to a dot product (scalar output) or matrix multiplication (matrix output).  This style of vector multiplication and division is standard in numerical array manipulations such as those in \texttt{numpy}, \texttt{IDL}, \texttt{matlab}, \emph{etc.}}

@@

A recreation of the unvarnished input spectrum---if desired---can be obtained by multiplying the continuum-flattened signal by the ``perturbed black body'', $\mathsf{B}(\bm{\lambda}_S)\odot \mathsf{P}(\bm{\lambda}_S)$, that symbolizes the black body modulated by continuum opacity or broad-band radiative transfer effects.  This smooth spectrum may be useful for applications that need to keep track of broad-band flux, such as low-resolution spectra, or regions with molecular band heads. The ``perturbed black body'' continuum model contains $n_{\mathrm{poly}}+1$ fixed-but-possibly tunable lookup parameters, plus the fixed input $T_{\mathrm{eff}}$.  For most practitioners, these terms serve as nuisance parameters and are perfunctorily discarded.

@@

The Voigt profile $\mathsf{V}(\lambda, \sigma_j, \gamma_j)$ can be computed in exact closed-form using the Voigt-Hjerting function \citep{1938ApJ....88..508H} as the real part of the complex Fadeeva function \citep[\emph{e.g.}][]{2011arXiv1106.0151Z}. Evaluation of the Fadeeva function can be computationally costly, so approximate forms may be desirable. Here we adopt the pseudo-Voigt approximation \citep{Ida:nt0146}.

@@

\subsection{Goodness of fit metric and gradients}
The model evaluated with its coarse initial values would have terrible performance: it would only vaguely resemble the synthetic spectral model, with up to $\pm 50\%$ undulations from the inexact assignment of widths, lineshapes, and amplitudes. Instead, we tune the parameters of the model, starting from these coarse initial values. This model has between $N_{\mathrm{lines}}\times 1$ and $N_{\mathrm{lines}}\times 4$ free parameters, depending on how many of the 4 physical line properties you wish to tune.  \edit2{The threshold-based line-finding algorithm performs very well at finding the line center positions, so in practice, we can fix the line center wavelengths.}  We, therefore, default to fitting three parameters per line, where the center wavelength is held fixed and the amplitude, width, and lineshape are allowed to vary. We minimize a scalar ``goodness-of-fit'' metric, \emph{aka} loss scalar $\mathcal{L}$, chosen as the mean squared error (MSE), which is proportional to $\chi^2$, the sum of the squares of the residual vector $\mathsf{R} \equiv \mathsf{S}-\mathsf{S}_{\rm clone}$ but has no notion of per-pixel noise since the precomputed synthetic spectrum has no uncertainty:

@@

\edit1{Geometrically, the $\grad \mathcal{L}$ Jacobian resembles a compass needle guiding the direction to the best fit. This compass needle resides in, say, 21,000-dimensional space, instead of our mere 3 dimensions of physical space.  Conventional gradient-free optimizers evaluate the Jacobian through approximate finite differences, which result in numerical instabilities.  Those instabilities accrue and break down at around the 300-dimension mark.  So non-gradient-based optimization can be pictured as having only a coarse needle---assembled from approximate finite differences---that begins to spin aimlessly after too many dimensions.}

@@

It is this ability to automatically compute Jacobians that sets \texttt{PyTorch} (and \texttt{JAX} and \texttt{TensorFlow}) apart.  These frameworks give exact Jacobians, instantaneously, for free (or cheap).  Without exact Jacobians, a conventional optimization step only obtains \emph{one} piece of information: how much the overall loss changed.  With exact Jacobians, we obtain $N_\mathrm{params}+1$ pieces of information for each evaluation of the forward model.  The power of gradient descent becomes transformative as the number of parameters grows into the tens of thousands, as is the case for \emph{blas\'e}.

@@

Forward modeling with tens of thousands of physics-informed parameters may seem like such a significant paradigm shift that it can feel too good to be true.  In this section, we introduce the non-negligible architectural design tradeoffs that arise when adopting an autodiff framework---such as \texttt{PyTorch}---for physics-based forward modeling.

First, we make a few tweaks to the implementation for numerical purposes. We enforce that all Gaussian and Lorentzian widths are positive by tuning the natural log of the widths, then exponentiating them before inclusion in Equation \ref{equation1}. Based on the minus sign in Equation \ref{equation1}, an amplitude $a_j$ could hypothetically take on either positive values (flux loss, absorption) or negative values (flux gain, emission).  For now, we focus on photospheric absorption lines---as opposed to, say, chromospheric emission lines---and therefore enforce all the amplitudes to be positive by tuning $\ln{a_j}$ and then exponentiating in the same strategy as above.  Emission lines---if desired---could be included with a mere sign flip to Equation \ref{equation1}.  Lines that can manifest either in absorption or emission could hypothetically relax the natural log pre-processing step for isolated lines.  We narrate only the absorption scenario moving forward.

@@

Sparse matrix methods generally support an operation known as \emph{coalescing}, which sums values with repeated indices. Each pixel may get computed about $\sim100$ times in this sparse implementation, which is about $70\times$ faster than each pixel getting computed $N_\mathrm{lines}\sim7000$ times in the dense approach.  Efficient algorithms for assembling and coalescing sparse matrices exist in \texttt{PyTorch}.  Some GPUs now support additional hardware acceleration of sparse matrices, providing even greater speedups.

@@

\begin{figure}[hbt!]
    \centering
    \includegraphics[width=1.0\columnwidth]{blase_clone_demo.png}
    \caption{PHOENIX spectrum cloned with \emph{blas\'e}.  This $T_{\mathrm{eff}}=4700\;$K,  $\log{g}=4.5$ solar metallicity model has 9,028 individual cloned spectral lines, each with 3 tuned parameters.  The pictured 50 \AA~ chunk contains 121 spectral lines and represents about 1$\%$ of the entire spectral bandwidth that was cloned.  Some flaws can be seen near the cores of deep lines or wings of broad lines.}
    \label{fig_cloned_spectrum_demo}
\end{figure}

@@

\begin{figure*}[hbt!]
    \centering
    \includegraphics[width=0.98\textwidth]{TelFit_clone_2panel.pdf}
    \caption{Pre-computed TelFit model cloned with \texttt{blase}.  \textbf{Before:} The top panel shows a forward model evaluated with coarsely initialized line-by-line properties, based on simple threshold-based peak-finding and crude estimation of the shape of the line.  \textbf{After:} The bottom panel shows the \texttt{blase} fine-tuned clone, with nearly identical spectral structure.}
    \label{fig_telluric_clone}
\end{figure*}

@@

We compute the residual $\mathsf{R}(\bm{\lambda}_S)$ of native-minus-cloned PHOENIX model, illustrated in the bottom panel of Figure \ref{fig_cloned_spectrum_demo}. We see an RMS residual of 1.2\%/pixel at native resolution.  The telluric clone shows a comparable level of performance.  These residuals tend to pile up in local symmetrically balanced clusters that get canceled out once convolved with coarser instrumental line profiles, so their overall effect at instrumental resolution is typically negligible: the clones are almost perfect.  We identify three main categories of cloning flaws that may not be negligible depending on the science application.

@@

Second, a conspicuous flaw occurs in the line cores of relatively narrow lines.  The cloned model tends to overestimate the flux at the core and underestimate the flux along the slopes of the lines.  \edit1{The residuals, therefore, exhibit a \texttt{W}-shape ringing artifact.  The scientific impact of this flaw will depend on the precision demands of the application, but it may not be as bad as it seems for many applications.  Recall that the cloning occurs at native spectral resolution, much greater than typical instrumental resolution.  The \texttt{W}-shape ringing residuals will tend to cancel each other out once the clone model undergoes convolution with an instrumental profile (Section \ref{jointModel}).  We ascribe this flaw to the adoption of approximate pseudo-Voigt profiles rather than exact Voigt profiles.  High-precision applications may wish to employ the exact Voigt profile, which we estimate to cost about 27$\times$ more than this approximate form.}

\edit1{The overprediction of line cores in narrow lines is only partly attributable to the pseudo-Voigt approximation.  Even exact Voigt profiles inadequately represent true atmospheric opacities. Typical stellar and substellar atmosphere models compute opacities for a stratified atmosphere consisting of tens or hundreds of locally isothermal, isobaric atmospheric layers.  Each of those layers may exhibit a local per-line opacity that matches a Voigt profile.  But the cumulative product of many slightly-different Voigt profiles is not exactly equal to any single Voigt profile.  So the \emph{blas\'e} approach of asserting lines as Voigt profiles can be best understood as a phenomenological model-- that lines tend to look Voigt-like, even if they are non-Voigt in detail.  In principle, \texttt{blas\'e} could be configured in a way to learn the requisite perturbations to the Voigt profile in a systematic manner, pooling information among lines.  Such an approach is not yet implemented and may be an enticing avenue for future research.}

@@

\begin{figure}[hbt!]
    \centering
    \includegraphics[width=0.48\textwidth]{blase_clone_demo_zoom.png}
    \caption{Zoom in of the region between the vertical gray bars in Figure \ref{fig_cloned_spectrum_demo}. The cloned model has 7 spectral lines describing the 400 pixels in this 4 \AA~ chunk.  The native PHOENIX pixel sampling can be seen as the boxy steps in both the native and cloned model.  }
    \label{fig_zoom_cloning_performance}
\end{figure}

@@

The cloned model already represents a useful intermediate product: the distillation of $N_s=335,849\times2$ pixel flux values and their wavelength coordinates into a more compact quartet of properties for a list of $<10,000$ spectral lines, a dimensionality reduction of $18\times$ for the cost of 1.2\%/pixel in accuracy.  So the process so far can be myopically viewed as a physics-informed compression algorithm.  But the cloned model serves as a mere stepping stone in our principal quest: the comparison of models to real data.

\subsection{Augmenting the stellar clone with radial velocity and rotational broadening }
Real stars possess two key extrinsic properties.  Rotational broadening $v\sin{i}$ and radial velocity $RV$ depend on the observer's viewing location. We follow \citet{czekala15} by emphasizing the qualifier \emph{extrinsic}, to distinguish between stellar \emph{intrinsic} properties, such as $T_{\mathrm{eff}}, \log{g},\mathrm{and\,} [\mathrm{Fe}/\mathrm{H}]$.  Intrinsic properties appear the same from any viewing location---at least for stars with isotropic surfaces---while extrinsic properties do not.  The distinction is important because the extrinsic terms act as simple convolutions and translations to the cloned spectrum, and can be treated after the cloning procedure.  We, therefore, define an augmented model, which we designate the ``extrinsic'' model, $\mathsf{S}_{\rm ext}$:

@@

The $\mathsf{S}_{\rm ext}$ spectrum is shown in Step 2 of Figure \ref{blase_flowchart}.  There is currently no equivalent post-processing of the telluric spectrum $\mathsf{T}$.  As mentioned previously, we assume the motions of Earth's atmosphere are much less than the desired stellar radial velocity precisions.  However, demanding Extreme Precision Radial Velocity (EPRV) exoplanet searches may need to consider minuscule systematic $RV$ shifts and broadening of the telluric templates, arising from the turbulent and bulk motions of the Earth's atmosphere.  A $\mathsf{T}_{\rm ext}$ could hypothetically be included in  \emph{blas\'e} to achieve these strenuous precision demands.

@@

Notice that the order of Steps 3 and 4 in Figure \ref{blase_flowchart} cannot be swapped.  Mathematically speaking, elementwise multiplication and convolution do not commute.  While the distinction may seem negligible, it matters at the level of precisions sought in EPRV applications (Suvrath Mahadevan, Arpita Roy, Sharon Xuesong Wang \emph{priv. comm.}).  Water vapor lines in our own atmosphere can ``beat'' with water vapor in the spectrum of, say, an M-dwarf atmosphere.  The systematic telluric mis-cancelation would imbue a Moir\'e pattern of residuals that is most acute for sources with sharp lines, namely low projected rotational broadening ($v\sin{i}\sim\frac{c}{2R}$).  The approach in \emph{blas\'e} may therefore unlock a level of telluric calibration that has evaded previous efforts.

As noted, typical data-pixel sampling $\bm{\lambda}_D$ is much coarser than the model pixel sampling $\bm{\lambda}_Z$.  We, therefore, resample the model to the data spectrum in the following way.  We evaluate the joint model $\mathsf{M}_{\rm joint}$ at all of the super-resolution wavelength coordinates and then compute the mean value of those pixels within the bounds of each coarse data pixel.  The resampling procedure is autodiff-aware: the same clusters of high-resolution coordinates map to the same data pixel coordinates, no matter what the $RV$ is.  The $RV$ only dictates what flux values are realized within those pixel bounds.

@@

It may appear desirable to simply optimize all of these parameters in a \emph{laissez-faire} manner, allowing them all to take on whatever value the data dictates.  Such a stratagem would overfit the data, resulting in unphysically perverse lineshapes that do not reflect the air of reality we aspire to impose on our synthetic spectral models.  Lines would haphazardly fit noise spikes, and conspire together to warp spectral shapes in unexpected ways.  This overfit model may suit some rare purposes.  But most of the time, we prefer to strike a better balance in the bias-variance tradeoff.

@@

The degree of regularization will control the extent of overfitting or underfitting. The most extreme regularization---the antithesis of the \emph{laissez-faire} scenario---would yield a model too rigid to respond to the data at all, yielding a model entirely unchanged from the cloned PHOENIX and \texttt{TelFit} models, the extreme end of underfitting.  So regularization constitutes the only hyperparameters worthy of tuning in \emph{blas\'e}.  The choice of how to set the regularization is problem-specific.  We default to the following choice. We fix all line parameters except for amplitude, which receives an L1 loss-- namely, we penalize the absolute value of departures from an amplitude's starting place:

@@

Figure \ref{fig_WASP69_demo} shows a portion of an HPF spectrum of WASP 69 centered on a region devoid of telluric lines.  This figure highlights a baseline case with no line-by-line fitting, simply conventional template matching: comparing HPF data of WASP 69 to the closest PHOENIX template, convolved and resampled to the HPF resolution and sampling.  Conspicuous residuals of $\pm10\%$ appear throughout the spectrum, with lines in the correct place, but with the amplitudes systematically biased.  These line residuals arise from bona-fide imperfections in the PHOENIX spectrum, with a minor contribution from the coarse sampling of the PHOENIX grid.

Figure \ref{fig_WASP69_transferred} shows the same data spectrum compared to a pixel-level model trained using the \emph{blas\'e} technique described in Section \ref{transferLearn}.  The model fit appears much better, with typical residuals approaching the photon noise of the data themselves.  The model is not perfect however, especially around line cores.  These residuals stem from a combination of causes.  First, our finite regularization restricts line amplitudes and widths from straying too far from their values.  This computational tug-of-war makes line cores land just short of the values they otherwise would have obtained in the absence of regularization.  Second, the spectral resolution kernel may get biased from the domineering model-misspecification of broad lines, setting up a slightly sub-par performance for all the other lines.

@@

Up to this point, we have presented the mechanics of  \emph{blas\'e}, and have shown that it applies to a wide range of applications including stars, substars, and Earth's atmosphere. The interpretation of the approach and its line-by-line outputs may carry different meaning to different practitioners.  Here we caricature some of these perspectives:
